Cover Your Basis: Comprehensive Data-Driven Characterization of the Binary Black Hole Population

### 2022-10-21 src/tex/conclusion.tex
update for arxiv

Non-parametric and data-driven statistical modeling methods have been put to use with great success across the ever-growing field of gravitational 
waves \citep{B_Farr_etal_2014,Littenberg_2015,Mandel_2016,Edwards_2018,Doctor_GPR,Edelman_2021,Vitale_2021,Tiwari_2021_a,Tiwari_2021_b,Edelman_2022ApJ,Tiwari_2022ApJ}. 
We presented a case study exploring how basis splines make for an especially powerful and efficient data driven method of characterizing the binary black hole population observed 
with gravitational waves, along with the associated open source software \textsc{GWInferno} (\url{https://git.ligo.org/bruce.edelman/gwinferno}) that implements the models 
described in this paper and performs hierarchical Bayesian inference with \textsc{NumPyro} and \textsc{Jax} \citep{pyro,numpyro,jax}. 
Our study paved the way as the first completely non-parametric compact object population study, employing data driven models for each of the hierarchically 
modeled population distributions. A complete understanding of the population properties of compact objects will help to advance poorly understood areas of stellar and 
nuclear astrophysics and provide a novel independent cosmological probe. With the coming influx of new data with the LVK's next observing run and the wide 
range of research avenues the compact object population will help to uncover, development of model-agnostic methods, such as the one we proposed here, will become necessary to efficiently make
sense of the data.

>>>>

Non-parametric and data-driven statistical modeling methods have been put to use with great success across the ever-growing field of gravitational 
waves \citep{B_Farr_etal_2014,Littenberg_2015,Mandel_2016,Edwards_2018,Doctor_GPR,Edelman_2021,Vitale_2021,Tiwari_2021_a,Tiwari_2021_b,Edelman_2022ApJ,Tiwari_2022ApJ}. 
We presented a case study exploring how basis splines make for an especially powerful and efficient data driven method of characterizing the binary black hole population observed 
with gravitational waves, along with the associated open source software \textsc{GWInferno} (\url{https://git.ligo.org/bruce.edelman/gwinferno}) that implements the models 
described in this paper and performs hierarchical Bayesian inference with \textsc{NumPyro} and \textsc{Jax} \citep{pyro,numpyro,jax}. 
Our study paves the way as the first completely non-parametric compact object population study, employing data driven models for each of the hierarchically 
modeled population distributions. A complete understanding of the population properties of compact objects will help to advance poorly understood areas of stellar and 
nuclear astrophysics and provide a novel independent cosmological probe. With the coming influx of new data with the LVK's next observing run, development of model-agnostic methods, such as the one we proposed here, will become necessary to efficiently make
sense of the vast amounts of data and to extract as much information as possible from the population.

### 2022-10-21 src/tex/intro.tex
update for arxiv

Observations of gravitational waves (GWs) from compact binary mergers are becoming a regular occurrence, 
producing a catalog of events that recently surpassed 90 such detections \citep{GWTC1,gwtc2,GWTC3}. As the catalog continues to grow, so does our understanding of the underlying astrophysical population of compact binaries \citep{o1o2_pop,o3a_pop,o3b_astro_dist}. 
Following numerous improvements to the detectors since the last observing run, the anticipated sensitivities for the upcoming fourth observing run of the LIGO-Virgo-KAGRA (LVK) collaboration suggest detection rates as high as once per \emph{day} \citep{aLIGO,aVIRGO,LVK_prospects,KAGRAProg}. With the formation history of these dense objects encoded in the details of their distribution \citep{Zevin_2017}, the likely doubling in size of the catalog with the next observing run could provide another leap in our understanding of compact binary astrophysics. 
Beyond formation physics, population-level inference of the compact binary catalog has also been shown to provide novel measurements of cosmological parameters \citep{Farr_2019HUB,gwtc3_cosmo,JoseSpectralSirens}, constrain modified gravitational wave propagation \citep{OkounkovaBirefringence,ModGWProp,ModGWProp2}, 
constrain a running Planck mass \citep{Lagos_runningPlanckMass}, search for evidence of ultralight bosons through superradiance \citep{Ng_Boson2021,GWTC2_superradiance_Ng}, 
constrain stellar nuclear reaction rates \citep{Farmer_2019,Farmer_2020}, look for primordial black holes \citep{Ng_2021,KenNgPBH2022}, 
and to constrain physics of neutron stars \citep{Golomb_EOS,LandryRead_NS_Masses2021}. Through a better understanding of the mass, spin, and redshift distributions of 
compact binaries that will come with the increased catalog size, one can probe a wide range of different physical phenomena with even greater fidelity.

The binary black hole (BBH) mass distribution was first found to have structure beyond a smooth power law with simpler parametric models, exhibiting a possible high mass truncation and either a break or a peak at $m_1\sim35-40\,\msun$ \citep{Fishbach_2017,Talbot_2018,o1o2_pop,o3a_pop}. Starting with moderately sized catalog, GWTC-2, more flexible models found signs of additional structure \citep{Tiwari_2021_b,Edelman_2022ApJ}. The evidence supporting these features, 
such as the peak at $m_1\sim10\,\msun$, has only grown after analyzing the latest catalog, GWTC-3, with the same models \citep{o3b_astro_dist,Tiwari_2022ApJ}. 
While this shows the usefulness of data-driven methods with the current relatively small catalog size, they will become more powerful with more observations. The canonical approach to constructing population models has been 
to use simple parametric descriptions (e.g., power laws, beta distributions) that aim to describe the data in the simplest way, employ astrophysically motivated priors where appropriate, then sequentially add 
complexities (e.g., Gaussian peaks) as the data demands. This simple approach was necessary when data was scarce, but as we move into the data-rich catalog era, this approach is already failing to scale.  More flexible and scalable methods, such as non-parametric modeling techniques, will be necessary to continue to extract the full information contained in the compact binary catalog. In contrast to parametric models, flexible and non-parametric models are data-driven and 
contribute little bias to functional form. Non-parametric or semi-parametric methods are particularly useful to search for unexpected features in the 
data, providing meaningful insight of these features that parametric models may fail to capture.

While we eventually hope to uncover hints of binary formation mechanisms in the mass spectrum of BBHs, the distribution of spin properties have been of particular interest.  
The measurement of spin properties of individual binaries often have large uncertainties, but the theorized formation channels are expected to produce distinctly 
different spin distributions \citep{Rodriguez_2016,Gerosa2018, Farr_BinnedSpin, Farr2017Nature, Zevin_2017}. Isolated (or field) formation scenarios predict component spins that are preferentially aligned with the binary's orbital angular momentum, although some small 
misalignment can occur depending on the nature of the supernova kicks as each star collapses to a compact object \citep{Zevin_2022, BaveraBBHSpin, BaveraMassTransfer}. Alternatively, dynamical formation in 
dense environments where many-body interactions between compact objects can result in binary formation and hardening (shrinking of binary orbits) should produce binaries with components' spins distributed isotropically \citep{Rodriguez_2016, Rodriguez_2019}. BBH spins have also been of controversial interest recently, with different parametric approaches to modeling 
the spin distribution coming to different conclusions. Studies have disagreed on the possible existence of a significant zero-spinning subpopulation, as well as the presence of 
significant spin misalignment (i.e. $\cos{\theta_i} < 0.0$) \citep{o3b_astro_dist,RouletGWTC2Pop,BuildBetterSpinModels,GWTC3MonashSpin,Callister_NoEvidence}. 
Another study recently showed that inferences of spin misalignment (or tilts) are sensitive to modeling choices and may not peak at perfectly aligned spins, as is often assumed \citep{spinitasyoulike}. While enlightening,
these recent efforts to improve BBH spin models continue to build sequentially on previous parametric descriptions \citep{BuildBetterSpinModels,Callister_NoEvidence,spinitasyoulike}. 
To ensure we are extracting the full detail the catalog has to offer, we extend our previous non-parametric modeling techniques to include spin magnitudes and tilts, as well as the binary mass ratio and redshift. 
%We present in this manuscript, both to help alleviate the current disputes in the literature and because there has been little work on flexible model development for the 
%spin distribution of BBHs beyond the aligned spin dimension thus far \citep{Tiwari_2021_a,Tiwari_2021_b}.

>>>>

Observations of gravitational waves (GWs) from compact binary mergers are becoming a regular occurrence, 
producing a catalog of events that recently surpassed 90 such detections \citep{GWTC1,gwtc2,GWTC3}. As the catalog continues to grow, so does our understanding of the underlying astrophysical population of compact binaries \citep{o1o2_pop,o3a_pop,o3b_astro_dist}. 
Following numerous improvements to the detectors since the last observing run, the anticipated sensitivities for the upcoming fourth observing run of the LIGO-Virgo-KAGRA (LVK) collaboration suggest detection rates as high as once per \emph{day} \citep{aLIGO,aVIRGO,LVK_prospects,KAGRAProg}. With the formation history of these dense objects encoded in the details of their distribution \citep{Rodriguez_2016,Farr2017Nature,Zevin_2017,Farr_BinnedSpin}, the likely doubling in size of the catalog with the next observing run could provide another leap in our understanding of compact binary astrophysics. 
Beyond formation physics, population-level inference of the compact binary catalog has also been shown to provide novel measurements of cosmological parameters \citep{Farr_2019HUB,gwtc3_cosmo,JoseSpectralSirens}, constrain modified gravitational wave propagation \citep{OkounkovaBirefringence,ModGWProp,ModGWProp2}, 
constrain a running Planck mass \citep{Lagos_runningPlanckMass}, search for evidence of ultralight bosons through superradiance \citep{Ng_Boson2021,GWTC2_superradiance_Ng}, 
constrain stellar nuclear reaction rates \citep{Farmer_2019,Farmer_2020}, look for primordial black holes \citep{Ng_2021,KenNgPBH2022}, 
and to constrain physics of neutron stars \citep{Golomb_EOS,LandryRead_NS_Masses2021}. Through a better understanding of the mass, spin, and redshift distributions of 
compact binaries that will come with the increased catalog size, one can probe a wide range of different physical phenomena with even greater fidelity.

The binary black hole (BBH) mass distribution was first found to have structure beyond a smooth power law with simpler parametric models, exhibiting a possible high mass truncation and either a break or a peak at $m_1\sim35-40\,\msun$ \citep{Fishbach_2017,Talbot_2018,o1o2_pop,o3a_pop}. Starting with the moderately sized catalog, GWTC-2, more flexible models found signs of additional structure \citep{Tiwari_2021_b,Edelman_2022ApJ}. The evidence supporting these features, 
such as the peak at $m_1\sim10\,\msun$, has only grown after analyzing the latest catalog, GWTC-3, with the same models \citep{o3b_astro_dist,Tiwari_2022ApJ}. 
While this shows the usefulness of data-driven methods with the current relatively small catalog size, they will become more powerful with more observations. The canonical approach to constructing population models has been 
to use simple parametric descriptions (e.g., power laws, beta distributions) that aim to describe the data in the simplest way, employ astrophysically motivated priors where appropriate, then sequentially add 
complexities (e.g., Gaussian peaks) as the data demands. This simple approach was necessary when data was scarce, but as we move into the data-rich catalog era, this approach is already failing to scale.  More flexible and scalable methods, such as the non-parametric modeling techniques presented in this manuscript, will be necessary to continue to extract the full information contained in the compact binary catalog. In contrast to parametric models, flexible and non-parametric models are data-driven and contribute little bias to functional form. They additionally are particularly useful to search for unexpected features in the 
data, providing meaningful insight into features that parametric models may fail to capture.

While we eventually hope to uncover hints of binary formation mechanisms in the mass spectrum of BBHs, the distribution of spin properties have been of particular interest.  
The measurement of spin properties of individual binaries often have large uncertainties, but the theorized formation channels are expected to produce distinctly 
different spin distributions \citep{Rodriguez_2016,Gerosa2018, Farr_BinnedSpin, Farr2017Nature, Zevin_2017}. Isolated (or field) formation scenarios predict component spins that are preferentially aligned with the binary's orbital angular momentum, although some small 
misalignment can occur depending on the nature of the supernova kicks as each star collapses to a compact object \citep{Zevin_2022, BaveraBBHSpin, BaveraMassTransfer}. Alternatively, dynamical formation in 
dense environments where many-body interactions between compact objects can result in binary formation and hardening (shrinking of binary orbits) should produce binaries with components' spins distributed isotropically \citep{Rodriguez_2016, Rodriguez_2019}. BBH spins have also been of controversial interest recently, with different parametric approaches to modeling 
the spin distribution coming to different conclusions. Studies have disagreed on the possible existence of a significant zero-spinning subpopulation, as well as the presence of 
significant spin misalignment (i.e. $\cos{\theta_i} < 0.0$) \citep{o3b_astro_dist,RouletGWTC2Pop,BuildBetterSpinModels,GWTC3MonashSpin,Callister_NoEvidence}. 
Another study recently showed that inferences of spin misalignment (or tilts) are sensitive to modeling choices and may not peak at perfectly aligned spins, as is often assumed \citep{spinitasyoulike}. While enlightening,
these recent efforts to improve BBH spin models continue to build sequentially on previous parametric descriptions \citep{BuildBetterSpinModels,Callister_NoEvidence,spinitasyoulike}. 
To ensure we are extracting the full detail the catalog has to offer, we extend our previous non-parametric modeling techniques to include spin magnitudes and tilts, as well as the binary mass ratio and redshift.

### 2022-10-21 src/tex/ms.tex
add zo's acknowledgements
add more citations -- remove \bruce{} flags
add ppc plot and some tables to try and make results presentation cleaner and more complete
go back to syw version==0.3.0 instead to fix clickable icons
update for arxiv

@@

\newcommand{\CIPlusMinus}[1]{{#1[median]^{+#1[error plus]}_{-#1[error minus]}}}
\newcommand{\CIBoundsBracket}[1]{{[#1[5th percentile], #1[95th percentile]]}}
\newcommand{\CIBoundsDash}[1]{{#1[5th percentile]\textendash#1[95th percentile]}}

@@

\usepackage{showyourwork}
\usepackage{bm}
\usepackage{amsmath}
\usepackage{appendix}
\input{macros}

@@

\begin{abstract}                 
We introduce the first complete non-parametric model for the astrophysical distribution of the binary black hole (BBH) population.  Constructed from basis splines, we use these models to conduct the most comprehensive data-driven investigation of the BBH population to date, including non-parametric modeling of BBH mass ratio, spin magnitudes and misalignments, and redshift distributions for the first time. Using the 69 confident BBHs in GWTC-3, we report the same features previously recovered 
with similarly flexible models of the mass distribution, most notably the peaks in merger rates at primary masses of ${\sim}10\msun$ and ${\sim}35\msun$.  We infer a distribution for primary spin misalignments that peaks away from alignment, supporting recent work that used a variety of parameterized models to come to the same conclusion. We find broad agreement with the 
previous inferences of the spin magnitude distribution: the majority of BBH spins are small ($a<0.5$), the distribution peaks at $a\sim0.2$, 
and there is mild support for a non-spinning subpopulation, which may be resolved with larger catalogs. 
Using our non-parametric models, we explore whether the data support spin distributions unique for each binary component, and 
find no strong evidence that the distributions differ. We conclude with a discussion of how non-parametric methods in 
gravitational-wave population inference are uniquely poised to complement to the parametric approach as we enter the 
data-rich era of gravitational-wave astronomy.

%When modeling the $\chi_\mathrm{eff}$ distribution with our model we find similar 
%conclusions as previously reported: the distribution is consistent with a non-isotropic distribution (that peaks at $\chi_\mathrm{eff}=\,$\result{$\CIPlusMinus{\macros[ChiEffective][chieff][PeakChiEff]}$}), 
%showing considerable support ($f_{\chi_\mathrm{eff}<0}=\,$\result{$\CIPlusMinus{\macros[ChiEffective][chieff][FracBelow0]}$}) for negative effective spins or misaligned systems.
\end{abstract}

\input{intro.tex}
\input{methods.tex}
\input{results.tex}
\input{conclusion.tex}

\section{Acknowledgements}\label{sec:acknowledments}
We thank Will Farr, Tom Callister, and Maya Fishbach for useful discussions during the preparation of this manuscript, 
along with helpful comments on an early draft by Tom Callister and Jaxen Godfrey.
This research has made use of data, software and/or web tools obtained from the Gravitational Wave Open Science Center 
(\url{https://www.gw-openscience.org/}), a service of LIGO Laboratory, the LIGO Scientific Collaboration and the Virgo Collaboration. 
The authors are grateful for computational resources provided by the LIGO Laboratory and supported by National Science Foundation Grants PHY-0757058 and PHY-0823459.  
This work benefited from access to the University of Oregon high performance computer, Talapas. This material is based upon work supported 
in part by the National Science Foundation under Grant PHY-1807046 and work supported by NSF's LIGO Laboratory which is a major facility 
fully funded by the National Science Foundation.
\software{
\textsc{Matplotlib}~\citep{Hunter:2007},
\textsc{NumPy}~\citep{harris2020array},
\textsc{SciPy}~\citep{2020SciPy-NMeth}
\textsc{AstroPy}~\citep{2018AJ....156..123A},
\textsc{Jax}~\citep{jax},
\textsc{NumPyro}~\citep{pyro,numpyro},
}
\bibliography{bib}{}
\bibliographystyle{aasjournal}

>>>>

\usepackage{showyourwork}
\usepackage{bm}
\usepackage{amsmath}
\usepackage{appendix}

\input{macros}
\newcommand{\CIPlusMinus}[1]{{#1[median]^{+#1[error plus]}_{-#1[error minus]}}}
\newcommand{\CIBoundsBracket}[1]{{[#1[5th percentile], #1[95th percentile]]}}
\newcommand{\CIBoundsDash}[1]{{#1[5th percentile]\textendash#1[95th percentile]}}

@@

@@

@@

\begin{abstract}                 
We introduce the first complete non-parametric model for the astrophysical distribution of the binary black hole (BBH) population.  Constructed from basis splines, we use these models to conduct the most comprehensive data-driven investigation of the BBH population to date, simultaneously fitting non-parametric models for the BBH mass ratio, spin magnitude and misalignment, and redshift distributions for the first time. Using the 69 confident BBHs in GWTC-3, we report the same features previously recovered 
with similarly flexible models of the mass distribution, most notably the peaks in merger rates at primary masses of ${\sim}10\msun$ and ${\sim}35\msun$. Our model reports a suppressed merger rate at low primary masses and a mass ratio distribution consistent with a power law. We infer a distribution for primary spin misalignments that peaks away from alignment, supporting recent work that used a variety of parameterized models to come to the same conclusion. We find broad agreement with the 
previous inferences of the spin magnitude distribution: the majority of BBH spins are small ($a<0.5$), the distribution peaks at $a\sim0.2$, 
and there is mild support for a non-spinning subpopulation, which may be resolved with larger catalogs. 
Using our non-parametric models, we explore whether the data support spin distributions unique for each binary component, and 
find no strong evidence that the distributions differ. With a modulated power law describing the BBH merger rate's evolution in redshift, 
we see hints of the rate evolution either flattening or decreasing at $z\sim0.2-0.5$, but the full distribution remains entirely consistent with a monotonically increasing power law. We conclude with a discussion of the astrophysical context of our new findings and how non-parametric methods in 
gravitational-wave population inference are uniquely poised to complement to the parametric approach as we enter the 
data-rich era of gravitational-wave astronomy.
\end{abstract}

\input{intro.tex}

\input{methods.tex}

\input{results.tex}

\input{conclusion.tex}

\section{Acknowledgements}\label{sec:acknowledments}
We thank Will Farr, Tom Callister, and Maya Fishbach for useful discussions during the preparation of this manuscript, 
along with helpful comments on an early draft by Tom Callister and Jaxen Godfrey. In the spirit of open source and reproducible science, this paper was written using the \textsc{showyourwork} package, published at \url{https://github.com/bruce-edelman/CoveringYourBasis} \citep{Luger2021}. Z.D. also acknowledges support from the CIERA Board
of Visitors Research Professorship. This research has made use of data, software and/or web tools obtained from the Gravitational Wave Open Science Center 
(\url{https://www.gw-openscience.org/}), a service of LIGO Laboratory, the LIGO Scientific Collaboration and the Virgo Collaboration. 
The authors are grateful for computational resources provided by the LIGO Laboratory and supported by National Science Foundation Grants PHY-0757058 and PHY-0823459.  
This work benefited from access to the University of Oregon high performance computer, Talapas. This material is based upon work supported 
in part by the National Science Foundation under Grant PHY-1807046 and work supported by NSF's LIGO Laboratory which is a major facility 
fully funded by the National Science Foundation.
\software{
\textsc{ShowYourWork}~\citep{Luger2021},
\textsc{Matplotlib}~\citep{Hunter:2007},
\textsc{NumPy}~\citep{harris2020array},
\textsc{SciPy}~\citep{2020SciPy-NMeth}
\textsc{AstroPy}~\citep{2018AJ....156..123A},
\textsc{Jax}~\citep{jax},
\textsc{NumPyro}~\citep{pyro,numpyro},
}
\bibliography{bib}{}
\bibliographystyle{aasjournal}
